Convective heat transport in stratified atmospheres at low and high Mach number

### 2016-08-31 tex/nusselt_paper.tex
moar updates, full 3 pages + spill over flushed out a bit

\section{Model \& Equations \label{section:model}}
We study a fluid whose equation of state is that of an ideal gas, $P = R^*\rho T$ and whose
initial stratification is polytropic.
The primary assumptions utilized in constructing a polytropic atmosphere are that the gravitational
acceleration and conductive heat flux are invariant throughout the depth of the atmosphere. In order
to satisfy the latter assumption, the thermal conductivity, $\kappa$ and temperature gradient
$\grad T_0$ are often taken as constants, such that, \bm{$F}_{\text{cond}} = -\kappa \grad T_0 = \text{constant}$.
Under these assumptions, solving the equation of hydrostatic equilibrium produces an atmosphere defined by
\begin{equation}
\begin{split}
\rho_0(z) &= \rho_{00}(z_0 - z)^m \\
T_0(z)    &= T_{00}(z_0 - z)
\label{eqn:polytrope}
\end{split}
\end{equation}
and $z$ increases upwards within the bounds $z =\{0, L_{z}\}$.
We specify the number of density scale heights the atmosphere spans, $n_\rho$ to determine $L_{z}$. Throughout
this study, we set $n_{\rho} = 3$ such that the density at the bottom of the atmosphere is larger
than that at the top of the atmosphere by roughly a factor of 20.
Thermodynamic variables are nondimensionalized at the top of the atmosphere as 
$P_0(L_z) = \rho_0(L_z) = T_0(L_z) = 1$, requiring $z_0 \equiv L_z + 1$ and $R^* = T_{00} = \rho_{00} = 1$.  
The polytropic index is set by the adiabatic index of a monatomic ideal gas, $\gamma = 5/3$ and the 
superadiabatic excess \cite{graham1975}, $\epsilon$, such that $m = (\gamma - 1)^{-1} - \epsilon = m_{ad} - \epsilon$ 
and the subsequent entropy gradient at the top of the atmosphere is $\grad S(L_z) = -\epsilon$,
the negative value of the superadiabatic excess.

Atmospheric diffusivities are set by the Rayleigh number and the Prandtl number.  We define the
non-dimensional Rayleigh number as
\begin{equation}
\text{Ra} = \frac{g L_z^3 (\Delta S_0 / c_P)}{\nu\chi},
\end{equation}
where $\Delta S_0$ is the entropy difference between the top and bottom of the atmosphere, 
$\nu$ is the kinematic viscosity (viscous diffusivity), and $\chi$ is the thermal dififusivity.  
The relationship between the thermal and viscous diffusivities is
set by the Prandtl number, Pr$ = \nu/\chi$.   We relate the dynamic viscosity, $\mu$ and the thermal conductivity,
$\kappa$, to their corresponding diffusivities such that 
$\nu \equiv \mu/\rho$ and $\chi \equiv \mu\/rho$.  As a result, $\text{Ra} \propto (\nu\chi)^{-1} \propto
\rho^2$, such that for our atmospheres with $n_{\rho} = 3$, the Rayleigh number increases by a factor of
approximately 400 from the top of the domain to the bottom of the domain.  Such a formulation leaves Pr
constant throughout the depth of the atmosphere, and in this paper we specifically study $\text{Pr} = 1$.

@@

\begin{figure}[t]
\includegraphics[width=3.4375in]{./figs/ma_v_eps.png}
\caption{Shown are characteristic mach numbers from IVPs spanning six decades of $\epsilon$, from
very low Mach number to near Mach one.  A very strong Ma $\propto \epsilon^{1/2}$ relation is clear. 
(need to actually do linear regression)
\label{fig:ma_v_eps} }
\end{figure}

@@

The efficiency of convection is defined by the Nusselt number.  While the Nusselt number is well-defined in \RB convection
as the amount of total flux divided by the steady-state background conductive flux 
\cite{johnston&doering2009, otero&all2002},
a well-defined Nusselt number is more elusive in stratified convection.  A traditional definition of the Nusselt
number in stratified convection is \cite{graham1975,hurlburt1984}
\begin{equation}
N \equiv \frac{F_{\text{conv, z}} + F_{\text{rad, z}} - F_A}{F_{\text{ref}} - F_A},
\label{eqn:nusselt}
\end{equation}
where $F_{\text{conv, z}}$ and $F_{\text{rad, z}}$ are the z-components of $\bm{F}_{\text{conv}}$ and $\bm{F}_{\text{rad}}$,
respectively.  $F_A$ is the adiabatic conductive flux, defined as $F_A = -\kappa \grad T_{\text{ad}}$.  For an
atmosphere in hydrostatic equilibrium, such as a polytrope, $\grad T_{\text{ad}} \equiv - g / c_{P}$, and thus
$F_A = \kappa g / c_{P}$.  $F_{\text{ref}} = \Delta T / L_z$ is the conductive flux of a linear profile connecting the upper
and lower plates, where $\Delta T = T_{u} - T_{\ell}$.

Such a definition of the Nusselt number is general to both stratified and \RB convection.  Convection works to
suppress entropy stratification and create isentropic atmospheres.  Under the Boussinesq approximation where
density variations are ignored, entropy stratification is directly proportional to temperature stratification,
such that $\grad S \rightarrow 0$ only when $\grad T \rightarrow 0$.  Thus, for \RB convection, 
$\grad T_{\text{ad}} \equiv 0$ and the familiar form of $N$ is retrieved.  In the case of stratified convection,
as $\epsilon \rightarrow m_{ad} + 1$, as in \cite{brandenburg2015}, $\grad P \propto g \rightarrow 0$ and
the resulting $\grad T_{\text{ad}} \rightarrow 0$.  In such a case, $F_A \rightarrow 0$ and the familiar
definition of the \RB nusselt number is appropriate to use. However, for any given values of $\epsilon$, including
the very small values used in this work, Eq. \ref{eqn:nusselt} is the proper non-dimensional definition
of the Nusselt number.

***WE USE DEDALUS, EXPLAIN WHAT IT ARE***

***EXPLAIN BOUNDARY CONDITIONS***

***EXPLAIN WHAT THE CHARACTERISTIC TIME SCALE, THE BUOYANCY TIME, IS***

\section{Results \label{section:results}}
We ran initial value problems for a few hundreds of buoyancy times past the convective transients from 
Rayleigh numbers around $R_{crit}$ up to Rayleigh numbers of $\approx 10^7 R_{crit}$.  While bulk thermodynamic
structures are similar between low and high $\epsilon$, high $\epsilon$ runs start to exhibit shock fronts
propagating away from downflow channels, such as those in Fig. \ref{fig:entropy_snapshots}, as reported in \cite{cattaneo&all1990}.
\begin{figure}[t]
\includegraphics[width=3.4375in]{./figs/snapshots_fig.png}
\caption{Two characteristic snapshots after about 200 buoyancy times are shown at $\epsilon=10^{-4}$ (top)
and $\epsilon=0.5$ (bottom) at Ra = $10^6$.  
\label{fig:entropy_snapshots} }
\end{figure}

Despite diferent thermodynamic structures, the fluxes look fairly similar at low and high mach number (sort of?
I don't think the ones I have are averaged over enough time, especially the eps=0.5 one, but I Pleiades is
a bit slammed right now).  See Fig \ref{fig:flux_profiles}

@@

At low Rayleigh number, diffusivities are high and the flows are very laminar.  Such flows often achieve a
steady state and have a well-defined Nusselt number which is independent of time.  However, as the Rayleigh
number increases, the flows become increasingly time-dependent.  Even steady structures such as solid
``rolls'' like those pictured in Fig. \ref{fig:entropy_snapshots} have highly time-dependent Nusselt numbers.
This is, in part, due to the fact that cold downdrafts floating to the bottom of the domain can be entrained
by upflows, or warm risen parcels can be entrainedi n the intense cold downdrafts.  Such events reverse the
preferred direction of flux in the system, and even let the Nusselt number become negative for short periods of
time.  See Fig. \ref{fig:nu_v_time}.  As a result, it is necessary to take a long time average of the fluxes
before calculating the Nusselt number at higher Rayleigh number.

@@

The evolution of the Nusselt number as the Rayleigh number is increased is shown for both high and low
$\epsilon$ in Fig. \ref{fig:nu_v_ra}.  Below convective onset, the Nusselt number is perfectly one. Just above
onset, there is a brief range of highly inflated scaling between $N$ and Ra.  From about $10R_{crit}$ to roughly
$10^{4-5}R_{crit}$, Ra and $N$ follow the relationship: (put a power law here)  Above about
$10^5 R_{crit}$, the Nusselt number flattens out as Ra is increased -- perhaps this is some Featherstone 2016
shennanigans.

\begin{figure}[b]
\includegraphics[width=3.4375in]{./figs/nu_v_ra.png}
\caption{Variation of the Nusselt number as a function of the Rayleigh number is shown.
\label{fig:nu_v_ra} }
\end{figure}

>>>>

\section{Model \& Equations \label{section:model}}
We study a fluid whose equation of state is that of an ideal gas, $P = R^*\rho T$ and whose
initial stratification is polytropic.
The primary assumptions utilized in constructing a polytropic atmosphere are that the gravitational
acceleration and conductive heat flux are invariant throughout the depth of the atmosphere. In order
to satisfy the latter assumption, the thermal conductivity, $\kappa$ and temperature gradient
$\grad T_0$ are often taken as constants, such that, \bm{$F}_{\text{cond}} = -\kappa \grad T_0 = \text{constant}$.
Under these assumptions, solving the equation of hydrostatic equilibrium produces an atmosphere defined by
\begin{equation}
\begin{split}
\rho_0(z) &= \rho_{00}(z_0 - z)^m \\
T_0(z)    &= T_{00}(z_0 - z)
\label{eqn:polytrope}
\end{split}
\end{equation}
and $z$ increases upwards within the bounds $z =\{0, L_{z}\}$.
We specify the number of density scale heights the atmosphere spans, $n_\rho$ to determine $L_{z}$. Throughout
this study, we set $n_{\rho} = 3$ such that the density at the bottom of the atmosphere is larger
than that at the top of the atmosphere by roughly a factor of 20.
Thermodynamic variables are nondimensionalized at the top of the atmosphere as 
$P_0(L_z) = \rho_0(L_z) = T_0(L_z) = 1$, requiring $z_0 \equiv L_z + 1$ and $R^* = T_{00} = \rho_{00} = 1$.  
The polytropic index is set by the adiabatic index of a monatomic ideal gas, $\gamma = 5/3$ and the 
superadiabatic excess \cite{graham1975}, $\epsilon$, such that $m = (\gamma - 1)^{-1} - \epsilon = m_{ad} - \epsilon$ 
and the subsequent entropy gradient at the top of the atmosphere is $\grad S(L_z) = -\epsilon$,
the negative value of the superadiabatic excess.  We argue that the characteristic timescale of such an
atmosphere is related to the characteristic atmospheric buoyancy time, $t_{\text{b}} = \sqrt{L_z/g\epsilon}$,
not the characteristic thermal diffusion time, $t_{\text{th}} = L_z^2/\chi$.

Atmospheric diffusivities are set by the Rayleigh number and the Prandtl number.  We define the
non-dimensional Rayleigh number as
\begin{equation}
\text{Ra} = \frac{g L_z^3 (\Delta S_0 / c_P)}{\nu\chi},
\end{equation}
where $\Delta S_0$ is the entropy difference between the top and bottom of the atmosphere, 
$\nu$ is the kinematic viscosity (viscous diffusivity), and $\chi$ is the thermal dififusivity.  
The relationship between the thermal and viscous diffusivities is
set by the Prandtl number, Pr$ = \nu/\chi$.   We relate the dynamic viscosity, $\mu$ and the thermal conductivity,
$\kappa$, to their corresponding diffusivities such that 
$\nu \equiv \mu/\rho$ and $\chi \equiv \mu/\rho$.  As a result, $\text{Ra} \propto (\nu\chi)^{-1} \propto
\rho^2$, such that for our atmospheres with $n_{\rho} = 3$, the Rayleigh number increases by a factor of
approximately 400 from the top of the domain to the bottom of the domain.  Such a formulation leaves Pr
constant throughout the depth of the atmosphere, and in this paper we specifically study $\text{Pr} = 1$.

@@

\begin{figure}[t]
\includegraphics[width=3.4375in]{./figs/ma_v_eps.png}
\caption{Shown are characteristic mach numbers from IVPs spanning six decades of $\epsilon$, from
very low Mach number to near Mach one.  A very strong Ma $\propto \epsilon^{1/2}$ relation is clear. 
\label{fig:ma_v_eps} }
\end{figure}

@@

The efficiency of convection is defined by the Nusselt number.  While the Nusselt number is well-defined in \RB convection
as the amount of total flux divided by the steady-state background conductive flux 
\cite{johnston&doering2009, otero&all2002},
a well-defined Nusselt number is more elusive in stratified convection.  A traditional definition of the Nusselt
number in stratified convection is \cite{graham1975,hurlburt&all1984}
\begin{equation}
N \equiv \frac{F_{\text{conv, z}} + F_{\text{rad, z}} - F_A}{F_{\text{ref}} - F_A},
\label{eqn:nusselt}
\end{equation}
where $F_{\text{conv, z}}$ and $F_{\text{rad, z}}$ are the z-components of $\bm{F}_{\text{conv}}$ and $\bm{F}_{\text{rad}}$,
respectively.  $F_A$ is the adiabatic conductive flux, defined as $F_A = -\kappa \grad T_{\text{ad}}$.  For an
atmosphere in hydrostatic equilibrium, such as a polytrope, $\grad T_{\text{ad}} \equiv - g / c_{P}$, and thus
$F_A = \kappa g / c_{P}$.  $F_{\text{ref}} = \Delta T / L_z$ is the conductive flux of a linear profile connecting the upper
and lower plates, where $\Delta T = T_{u} - T_{\ell}$.

Such a definition of the Nusselt number is general to both stratified and \RB convection.  Convection works to
suppress entropy stratification and create isentropic atmospheres.  Under the Boussinesq approximation where
density variations are ignored, entropy stratification is directly proportional to temperature stratification,
such that $\grad S \rightarrow 0$ only when $\grad T \rightarrow 0$.  Thus, for \RB convection, 
$\grad T_{\text{ad}} \equiv 0$ and the familiar form of $N$ is retrieved.  In the case of stratified convection,
as $\epsilon \rightarrow m_{ad} + 1$, as in \cite{brandenburg&all2005}, $\grad P \propto g \rightarrow 0$ and
the resulting $\grad T_{\text{ad}} \rightarrow 0$.  In such a case, $F_A \rightarrow 0$ and the familiar
definition of the \RB nusselt number is appropriate to use. However, for any given values of $\epsilon$, including
the very small values used in this work, Eq. \ref{eqn:nusselt} is the proper non-dimensional definition
of the Nusselt number.

We utilize the novel Dedalus pseudospectral framework (cite website?) to time-evolve Eqs. 
\ref{eqn:continuity_eqn}-\ref{eqn:energy_eqn}.  We use a dealiased Chebyshev (vertical)
and Fourier (horizontal) grid in which the
physical grid dimensions are 3/2 the size of the coefficient grid.  Our coefficient grid sizes range from
64x256 grid points (low Ra) to 768x3072 grid points (Ra = $10^{7.66}$).  The atmosphere is bounded above
and below by impenetrable, stress free, fixed temperature boundary conditions such that
\begin{equation}
w = \partial_z u = T_1 = 0
\end{equation}
at the boundaries.  Theh domain is horizontally periodic.

\section{Results \label{section:results}}

We ran initial value problems for a few hundreds of buoyancy times past the convective transients from 
Rayleigh numbers around $R_{crit}$ up to Rayleigh numbers of $\approx 10^7 R_{crit}$ (BOOTSTRAPPING?).  At
high values of $\epsilon$, shock systems form in the upper atmosphere near downflow lanes 
(see e.g. Fig. \ref{fig:entropy_snapshots}b) and propagate towards upflow lanes.  Such systems were reported in
both two \cite{cattaneo&all1990} and three \cite{malagoli&all1990} dimensional polytropic simulations previously.
Low mach number flows, such as those on an $\epsilon = 10^{-4}$ atmosphere (e.g. Fig. \ref{fig:entropy_snapshots}a)
have similar bulk thermodynamic structure but lack the complicating dynamics of shock heating.

\begin{figure}[t]
\includegraphics[width=3.4375in]{./figs/snapshots_fig.png}
\caption{Two characteristic snapshots after about 200 buoyancy times are shown at $\epsilon=10^{-4}$ (top)
and $\epsilon=0.5$ (bottom) at Ra = $10^6$.  
\label{fig:entropy_snapshots} }
\end{figure}

Despite diferent thermodynamic structures, the properly normalized \emph{sum} of the fluxes for a given
Ra is similar.  See, for example, Fig. \ref{fig:flux_profiles} to see time-averaged horizontal profiles of
the vertical system fluxes.  At high $\epsilon$ and Ra, $\bm{F}_{\text{enth}}$ and $\bm{F}_{\text{KE}}$
exhibit two local maxima over the depth of the domain, whereas at low $\epsilon$ and low Ma, there is only
one local maximum near the bottom of the domain.  Viscous fluxes become important near the bottom boundary
layer, and radiative flux is nearly zero everywhere in the domain except for near the boundaries where the
fixed-temperature boundary conditions create large deviations in $\grad T$ away from $\grad T_{\text{ad}}$.

@@

At low Rayleigh number, diffusivities are high and the flows are very laminar.  Such flows often achieve a
steady state and have a well-defined Nusselt number which is independent of time.  However, as the Rayleigh
number increases, the flows become increasingly time-dependent.  Even steady structures such as solid
``rolls'' like those pictured in Fig. \ref{fig:entropy_snapshots} have highly time-dependent Nusselt numbers.
This is, in part, due to the fact that cold downdrafts floating to the bottom of the domain can be entrained
by upflows, or warm risen parcels can be entrainedi n the intense cold downdrafts.  Such events reverse the
preferred direction of flux in the system, and even let the Nusselt number become negative for short periods of
time.  See Fig. \ref{fig:nu_v_time}.  As a result, it is necessary to take a long time average of the fluxes
before calculating the Nusselt number at higher Rayleigh number. (IS THIS USEFUL?)

@@

The evolution of the Nusselt number as the Rayleigh number is increased is shown for both high and low
$\epsilon$ in Fig. \ref{fig:nu_v_ra}.  Below convective onset, the Nusselt number is perfectly one. Just above
onset, there is a brief range of highly inflated scaling between $N$ and Ra.

\begin{figure}[t]
\includegraphics[width=3.4375in]{./figs/nu_v_ra.png}
\caption{Variation of the Nusselt number as a function of the Rayleigh number is shown.
\label{fig:nu_v_ra} }
\end{figure}
